---
title: "Bitcoin vs. Gold: Safe Haven Showdown (2015-2025)"
author: "Alexander G"
date: "2025-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Introduction
This project compares Bitcoin and gold as safe haven assets over a 10-year period (May 2015 to May 2025). I analyze their cumulative returns, volatility, and performance during market downturns to determine which asset is a better safe haven.

### Data Sources
*Bitcoin Prices: Sourced from Blockchain.com API (historical market price data).

*Gold Prices: Sourced from Alpha Vantage API using the GLD ETF (SPDR Gold Shares) as a proxy for gold prices.

*Data was cleaned in Google Sheets and processed in BigQuery to calculate weekly returns.

## Analysis

```{r load-data, echo=FALSE}

data <- read.csv("bitcoin_vs_gold_analyzed_10years.csv")
data$Date <- as.Date(data$Date)

```

### Cumulative Returns
The plot below shows the cumulative returns of Bitcoin and gold over the 10-year period.

```{r cumulative-returns-plot, echo=FALSE, fig.width=10, fig.height=6}
# Instead of loading the entire scales package, use :: notation to access only the function we need
# This avoids the namespace conflicts

# Option 1: Using pseudo_log10_trans (best approach)
ggplot(data, aes(x = Date)) +
  geom_line(aes(y = Bitcoin_Cumulative_Return, color = "Bitcoin"), linewidth = 1) +
  geom_line(aes(y = Gold_Cumulative_Return, color = "Gold"), linewidth = 1) +
  labs(title = "Cumulative Returns: Bitcoin vs. Gold (2015-2025)",
       x = "Date",
       y = "Cumulative Return (%) - Pseudo-Log Scale",
       color = "Asset") +
  scale_color_manual(values = c("Bitcoin" = "orange", "Gold" = "gold")) +
  scale_y_continuous(trans = scales::pseudo_log_trans(base = 10)) +
  theme_minimal() +
  theme(legend.position = "top")

```


You can also explore an interactive version of this analysis in the [Tableau Dashboard](https://public.tableau.com/views/Bitcoinvs_GoldSafeHavenAnalysis2015-2025/BitcoinvsGoldDashboard?:language=en-US&:sid=&:redirect=auth&:display_count=n&:origin=viz_share_link).


### Key Metrics

```{r metrics, echo=FALSE}

weeks_per_year <- 52
bitcoin_volatility <- sd(data$Bitcoin_Weekly_Return / 100, na.rm = TRUE) * sqrt(weeks_per_year)
gold_volatility <- sd(data$Gold_Weekly_Return / 100, na.rm = TRUE) * sqrt(weeks_per_year)

downturn_threshold <- -10
downturns <- data[data$Bitcoin_Weekly_Return < downturn_threshold, ]
bitcoin_downturn_avg <- mean(downturns$Bitcoin_Weekly_Return, na.rm = TRUE)
gold_downturn_avg <- mean(downturns$Gold_Weekly_Return, na.rm = TRUE)

cat("**Annualized Volatility**\n")
cat("- Bitcoin:", round(bitcoin_volatility, 2), "%\n")
cat("- Gold:", round(gold_volatility, 2), "%\n\n")
cat("**Average Return During Bitcoin Downturns (>10% Drop)**\n")
cat("- Bitcoin:", round(bitcoin_downturn_avg, 2), "%\n")
cat("- Gold:", round(gold_downturn_avg, 2), "%\n")

```

## Conclusion
*Cumulative Returns: Bitcoin likely outperformed gold significantly, but with higher volatility.

*Volatility: Gold exhibits lower volatility, making it a more stable asset.

*Safe Haven Performance: During Bitcoin downturns, gold typically performs better (e.g., positive or less negative returns), supporting its role as a safe haven.

*Recommendation: Gold appears to be the better safe haven due to its stability and performance during market stress, though Bitcoin may appeal to risk-tolerant investors seeking higher returns.




